---
title: "Tips for taking R apart"
output: ioslides_presentation
---

## 1. Evaluate the full line and see what the output looks like

```{r, echo = FALSE}
available.rows <- 1:27
available.rowseats <- LETTERS[1:6]
available.seats <- expand.grid(available.rows, available.rowseats)
available.seats <- apply(available.seats, 1, paste, collapse = "")
available.seats <- gsub(" ", "", available.seats) ## don't worry too much about this,
                                                  ## it just seems that there were 
                                                  ## some extra spaces at the start of
                                                  ## some seat names, so I'm cleaning
                                                  ## that up.

manifest <- data.frame(passenger = 1:(6 * 27),
                       assigned.seat = available.seats[sample(1:length(available.seats))], 
                       stringsAsFactors = FALSE)
manifest$row <-  as.numeric(gsub("[A-Z]", "", manifest$assigned.seat))
manifest$seat <- gsub("[0-9]", "", manifest$assigned.seat)
```

Evaluate: 

```{r}
seat.availability <- factor(rep("Empty", length.out = nrow(manifest)),
                            levels = c("Empty", "Taken"))
```

Then check:

```{r}
head(seat.availability)
length(seat.availability)
```

## 2. Check out any objects within the call

```{r}
seat.availability <- factor(rep("Empty", length.out = nrow(manifest)),
                            levels = c("Empty", "Taken"))
```

Check out: 

```{r}
head(manifest, 3)
```

## 3. Evaluate calls within calls

```{r}
seat.availability <- factor(rep("Empty", length.out = nrow(manifest)),
                            levels = c("Empty", "Taken"))
```

For this call, check out:
```{r}
nrow(manifest)
head(rep("Empty", length.out = nrow(manifest)))
c("Empty", "Taken")
```

## 4. Use "?" to find out about functions

```{r}
seat.availability <- factor(rep("Empty", length.out = nrow(manifest)),
                            levels = c("Empty", "Taken"))
```

For this call, check out:
```{r, eval=FALSE}
?nrow
?rep
?factor
```

Also, try googling "basic R example [function name]"

## 5. If you're looking at a loop, assign i to the first value of the loop then check out the code in the loop

```{r, eval=FALSE}
j <- 0
for(i in 1:3){
        j <- j + i
}
```

For this loop, check out:
```{r}
j <- 0

i <- 1
j <- i + 1
j
```